Relaxed Weak Accelerated Proximal Gradient Method: a Unified Framework for Nesterov's Accelerations

### 2025-09-15 Notebook LaTEX/R-WAPG/presentation-long.tex
>w<, changes

% SLIDE INFORMATION ============================================================
\author[Hongda Li]{Hongda Li\\ \vspace{1cm} \tiny{Joint work with Xianfu Wang}}
\title[Thesis Proposal Talk]{
    Relaxed Weak Accelerated Proximal Gradient Method: a Unified Framework for Nesterov's Accelerations
}
% \newcommand{\email}{lalala@lala.la}
\institute[UBCO]{
    University of British Columbia Okanagan
}
\date{\today}
\subject{Nesterov's acceleration and its applications}

@@

                \item When it's equal, as in the case of Nesterov's 1983, the convergence rate is the best. 
                \item It's showed by Chambolle and Dossal in 2015 \cite{chambolle_convergence_2015} that the choice $(k + a -1)/a$ has weak convergence of iterates $(x_k)_{k \ge 0}$ for all $a > 2$, under the same assumption as before. 

@@

            Fixing the momentum sequence gives an optimal convergence rate of the algorithm for the class of $F:\RR^n \rightarrow \RR$ that has 

@@

                \item and $\mu \ge 0$ strongly convexity. 

@@

            Available in Nesterov's book \cite{nesterov_lectures_2018} and Beck's book \cite{beck_first-order_2017} showed that if the sequence $(\theta_k)_{k\ge 0}$ has: 

@@

                \item A linear convergence rate of $\mathcal O((1 - \sqrt{\mu/L})^k)$ for $(F(x_k) - F(x^+))_{k \ge 1}$. 

@@

                Let the ambient space be $\RR^n$ equipped with Euclidean inner product and norm. 

@@

        \begin{frame}{Our theoretical contributions, statement of results}
            Here are the promises: 

@@

                \item We proposed a unified framework which we call ``Relaxed Weak Accelerated Proximal Gradient (R-WAPG)" that unifies the convergence theories for functions that are strongly convex or not necessarily strongly convex. 
                \item We demonstrate that the R-WAPG framework is equivalent to various representations of the Nesterov's type accelerated proximal gradient methods with Euclidean geometry in the literature. 

@@

        \begin{frame}{Summarized results of our theoretical contributions}

@@

            Suppose that $\alpha_0 \in (0, 1]$ and for all $k \ge 1$, $\alpha_k \in (\mu/L, 1)$ and define $(\rho_k)_{k\ge0 }$ by: 

@@

            We can show Nesterov's type accelerated proximal gradient algorithm (or equivalently, FISTA with $g \equiv 0$) generates $(F(x_k) - F^+)_{k\ge 1}$ has an upper bound of: 

@@

        \begin{frame}{Our convergence theory unifies variants in the literatures}

@@

            Our theories stated that, the convergence of R-WAPG algorithm is ultimately described by the sequence $(\alpha_k)_{k \ge 0}$.

@@

                \par
                It's completely parameter free by equipping it with Armijo line search. 

@@

                Let $(\alpha_k)_{k \ge 0}$ be a sequence such that it satisfies: 

@@

                    & \alpha_k \in (\mu/L, 1] \quad (\forall k \ge 1). 

@@

                Define $(\rho_k)_{k \ge 0}$ by 

@@

@@

            After six pages of dense proof in the paper, we can deduce the following theorem: 

@@

                Let vector sequence $(y_k, v_{k}, x_{k})_{k \ge 1}$ and R-WAPG sequences $\alpha_k, \rho_k$ be given by Definition \ref{def:wapg}.
                % Let $\epsilon_k, y_k$ be given by Assumption \ref{ass:regret-gap}.
                % Define $R_1 = 0$ and suppose that for $k = 1, 2, \ldots, N$, we have $R_k$ recursively given by:
                % \begin{align*}
                %     R_{k + 1}
                %     :=
                %     \frac{1}{2}\left(
                %         L^{-1} - \frac{\alpha_k^2}{\hat \gamma_{k + 1}}
                %     \right)\Vert g_k\Vert^2
                %     +
                %     (1 - \alpha_k)
                %     \left(
                %         \epsilon_k + R_k +
                %         \frac{\mu\alpha_k\gamma_k}{2\hat \gamma_{k + 1}}
                %         \Vert v_k - y_k\Vert^2
                %     \right).
                % \end{align*}

@@

\section{Our paper is on arxiv and references}
    \begin{frame}{Thanks for the participations. Merci pour les participations.}
        Our paper can be found on arxiv using the following QR code: 
        \begin{figure}
            \centering
            \includegraphics[width=15em]{assets/paper-qrcode.png}
        \end{figure}
        Questions?
    \end{frame}
    \begin{frame}{References}        
        \bibliography{references/R-WAPG.bib}
    \end{frame}

>>>>

\AtBeginSection[]{
  \begin{frame}
  \vfill
  \centering
  \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
    \usebeamerfont{title}\insertsectionhead\par%
  \end{beamercolorbox}
  \vfill
  \end{frame}
}

% SLIDE INFORMATION ============================================================

\author[Hongda Li]{Hongda Li}

\title{
    Relaxed Weak Accelerated Proximal Gradient Method: a Unified Framework for Nesterov's Accelerations
}
% \newcommand{\email}{lalala@lala.la}
\institute[UBCO]{
    University of British Columbia Okanagan
}
\date[\today]{\today \\ \vspace{1cm} \tiny{Joint work with Shawn/Xianfu Wang}}
\subject{Nesterov's acceleration and its applications}

@@

                \item When it's equal, as in the case of Nesterov's 1983, it gives the best convergence rate for all convex functions. 
                \item Showed by Chambolle and Dossal in 2015 \cite{chambolle_convergence_2015}, if we set the sequence $t_k = (k + a -1)/a$, then the iterates $(x_k)_{k \ge 0}$ has weak convergence in Hilbert space for all $a > 2$. 

@@

            Assume $F:\RR^n \rightarrow \RR$ that has:

@@

                \item Strongly convex with $\mu > 0$. 

@@

            In Nesterov's book \cite{nesterov_lectures_2018} and Beck's book \cite{beck_first-order_2017}, they showed that if $(\theta_k)_{k\ge 0}$ is:

@@

                \item A linear convergence rate $\mathcal O((1 - \sqrt{\mu/L})^k)$ for $(F(x_k) - F(x^+))_{k \ge 1}$. 

@@

                Let the ambient space be $\RR^n$, equipped with Euclidean inner product and norm. 

@@

        \begin{frame}{Our theoretical contributions, and results}

@@

                \item We proposed a unified framework which we call ``Relaxed Weak Accelerated Proximal Gradient (R-WAPG)" that unifies the convergence theories for $f$ that is strongly convex and, convex but not necessarily strongly convex. 
                \item We demonstrate that the R-WAPG framework is equivalent to various other representations of the Nesterov's type accelerated proximal gradient in the literatures. 

@@

        \begin{frame}{Theoretical results summary}

@@

            Suppose that $\alpha_0 \in (0, 1]$ and for all $k \ge 1$, $\alpha_k \in (\mu/L, 1)$ and define $(\rho_k)_{k\ge0 }$ to be: 

@@

            We can show Nesterov's type accelerated proximal gradient algorithm (or equivalently, FISTA with $g \equiv 0$) generates $(F(x_k) - F^+)_{k\ge 1}$ that has an upper bound of: 

@@

        \begin{frame}{Unifying variants of FISTA in the literatures}

@@

            Our theories sugguest that, the upperbound for function value in the R-WAPG algorithm is ultimately described by the sequence $(\alpha_k)_{k \ge 0}$.

@@

@@

                Let $(\alpha_k)_{k \ge 0}$ be such that: 

@@

                    & \alpha_k \in (\mu/L, 1) \quad (\forall k \ge 1). 

@@

                Define $(\rho_k)_{k \ge 0}$

@@

        \begin{frame}{Proximal gradient, gradient mapping}
            \begin{definition}[The proximal gradient operator]
                Define for all $x\in \RR^n$: 
                    T_L(y) 
                    &:= \argmin_{x \in \RR^n} \left\lbrace
                        g(x) + \langle \nabla f(y), x\rangle + L/2\Vert x - y\Vert^2
                    \right\rbrace 
                    &= \left(I + L^{-1}\partial g\right)^{-1}
                    \left(I - L^{-1}\nabla f\right)(y),
                    \mathcal G_L(y)
                    &:= L(I - T_L)y.
            Note, the $I$ here is the identity operator in $\RR^n$. 
            $L > 0$ is the Lipschitz smoothness parameter for $f$. 

@@

            After six pages of math in the paper, we deduced the following theorem: 

@@

                Let $(\alpha_k)_{k \ge 0}, (\rho_k)_{k \ge 0}$ be R-WAPG sequences.
                Let vector sequences $(y_k, v_{k}, x_{k})_{k \ge 1}$ given by Definition \ref{def:wapg}. 

@@

\section{references and draft on arxiv}
    \begin{frame}{Thanks for the participations. Merci pour les participations.}
        Our paper can be found on arxiv using the following QR code: 
        \begin{figure}
            \centering
            \includegraphics[width=15em]{assets/paper-qrcode.png}
        \end{figure}
        Questions?
    \end{frame}
    \begin{frame}{References}        
        \bibliography{references/R-WAPG.bib}
    \end{frame}

### 2025-09-15 Notebook LaTEX/R-WAPG/presentation-short.tex
>w<, changes

\AtBeginSection[]{
  \begin{frame}
  \vfill
  \centering
  \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
    \usebeamerfont{title}\insertsectionhead\par%
  \end{beamercolorbox}
  \vfill
  \end{frame}
}

@@

        \begin{frame}{Chambolle, Dossal}
            Generalized by Beck \cite{beck_fast_2009-1} in 2009 is the FISTA algorithm for additive smooth and non-smooth objective. 
            It's available in their proof that the sequence $(t_k)_{k}$ can be relaxed: 

@@

                \item When it's equal, as in the case of Nesterov's 1983, it gives the best convergence rate for all convex functions. 
                \item Showed by Chambolle and Dossal in 2015 \cite{chambolle_convergence_2015}, if we set the sequence $t_k = (k + a -1)/a$, then the iterates $(x_k)_{k \ge 0}$ has weak convergence in Hilbert space for all $a > 2$. 
        \begin{frame}{V-FISTA and strongly convex variants}
            Assume $F:\RR^n \rightarrow \RR$ that has:
                \item $L$ smooth gradient, 
                \item Strongly convex with $\mu > 0$. 
            In Nesterov's book \cite{nesterov_lectures_2018} and Beck's book \cite{beck_first-order_2017}, they showed that if $(\theta_k)_{k\ge 0}$ is:

@@

            Then the algorithm has
                \item A linear convergence rate $\mathcal O((1 - \sqrt{\mu/L})^k)$ for $(F(x_k) - F(x^+))_{k \ge 1}$. 

@@

@@

        \begin{frame}{Our theoretical contributions, and results}
                \item We relaxed the traditional choices of the sequence $(\theta_k)_{k \ge 1}$ to full extend to show that an upper bound for the sequence $(F(x_k) - F^+)_{k \ge 1}$ exists. 
                \item We proposed a unified framework which we call ``Relaxed Weak Accelerated Proximal Gradient (R-WAPG)" that unifies the convergence theories for $f$ that is strongly convex and, convex but not necessarily strongly convex. 
                \item We demonstrate that the R-WAPG framework is equivalent to various other representations of the Nesterov's type accelerated proximal gradient in the literatures. 

@@

            \textbf{Taking this idea further, this is what we did in practice}: 

@@

                We made the Free R-WAPG algorithm estimates the strong convexity $\mu$ constant through the Bregman Divergence at two successive iterates while the algorithm is running and put it into the sequence to run the algorithm.

@@

@@

                The algorithm generates a sequence of vector $(y_k, x_{k + 1}, v_{k + 1})_{k \ge 1}$ for $k\ge 1$ by the procedures:
                    For $k=1, 2, 3, \ldots$

@@

                        \hat \gamma_{k + 1} & \defeq (1 - \alpha_k)\gamma_k + \mu \alpha_k = L\alpha_k^2,

@@

                        y_k &\defeq

@@

                        g_k &\defeq \mathcal G_L (y_k),

@@

                        v_{k + 1} &\defeq

@@

                        x_{k + 1} &\defeq T_L y_k.

@@

            To use this theorem for the convergence of existing variants of Accelerated Proximal Gradient in the literature, we need alternative representations of R-WAPG that make it similar to what appears in the literature. 

@@

        \begin{frame}{The intermediate form}
            \begin{definition}[R-WAPG intermediate form]\label{def:r-wapg-intermediate}

@@

                Assume $\mu < L$ and let $(\alpha_k)_{k \ge 0}, (\rho_k)_{k \ge 0}$ given by Definition \ref{def:rwapg-seq}.
                Initialize any $x_1, v_1$ in $\RR^n$.
                For $k \ge 1$, the algorithm generates sequence of vector iterates $(y_{k}, v_{k + 1}, x_{k + 1})_{k \ge 1}$ by the procedures:
                    For $k = 1, 2, \ldots$
                        & y_{k} \defeq
                        \right),
                        & x_{k + 1} \defeq
                        & v_{k + 1} \defeq
                        \right) - \frac{1}{L\alpha_{k}}\mathcal G_L (y_k).
            The closest algorithm that fits the form but with $\mu = 0$ is Chapter 12 of in Ryu and Yin's Book \cite{ryu_large-scale_2022}, right after Theorem 17. 
        \begin{frame}{Intermediate form equivalency}
            \begin{proposition}[R-WAPG is equivalent to intermediate form]\label{prop:wapg-first-equivalent-repr}
                then the iterates can be expressed without $(\gamma_k)_{k \ge1},(\hat \gamma_k)_{k \ge 2}$,  namely, for all $k\ge 1$,

@@

                    & y_{k} =
                    \right), 
                    & v_{k + 1} =
                    \right) - \frac{1}{L\alpha_{k}}\mathcal G_L (y_k).
                    % \label{eqn:rwapg-first-equiv-form-eqn-2}
        \begin{frame}{Similar triangle form}
            \begin{proposition}[R-WAPG is equivalent to similar triangle form]\label{prop:wagp-st-form}
                Let iterates $(y_k, x_{k}, v_{k})_{k \ge 1}$ and sequence $(\alpha_k, \rho_k)_{k \ge 0}$ be given by Definition \ref{def:r-wapg-intermediate}.
                Then for all $k \ge 1$, iterates $y_k, x_{k + 1}, v_{k + 1}$
                satisfy:

@@

                    \right),
                    \label{eqn:rwapg-st-form-eqn-1}

@@

                    x_{k + 1} &=

@@

                    v_{k + 1} &= x_{k + 1} + (\alpha_k^{-1} - 1)(x_{k + 1} - x_k).
                    \label{eqn:rwapg-st-form-eqn-3}

@@

            Combining both of these representations, we can draw a similar triangle. 
        \begin{frame}{Similar triangle form illustration Part I}
            To understand the name ``similar triangle form" for it in the literature, we draw it out. 
            Write $\tau_k = \frac{L - L\alpha_k}{L\alpha_k - \mu}$ so $y_k = (1 + \tau_k)^{-1}(v_k + \tau_k x_k)$ from the similar triangle form. 
            The R-WAPG sequence has $\alpha_k \in (\mu/L, 1)$ for all $k \ge 1$, so it has: 
                \begin{figure}
                    \includegraphics[width=15em]{assets/st_part1.png}
        \begin{frame}{Similar triangle form illustration part II}
            Then $x_{k + 1} = y_k - L^{-1}\mathcal G_L(y_k)$ hence it has 
                \begin{figure}
                    \includegraphics[width=15em]{assets/st_part2.png}
        \begin{frame}{Similar triangle form illustration part III}
            Recall that $v_{k + 1} = x_{k + 1} + (\alpha_k^{-1} - 1)(x_{k + 1} - x_k)$, so $v_{k + 1}$ is on the line defined by $x_{k + 1} - x_k$, and it looks like: 
                \begin{figure}
                    \includegraphics[width=15em]{assets/st_part3.png}
            Let $\xi_k = \mu(L\alpha_k - \mu)^{-1}$, the intermediate form it has $v_{k+1} = (1 + \xi_k)^{-1}(v_k + \xi_k y_k) - L^{-1}\alpha_k^{-1} \mathcal G_L(y_k)$, hence the gray line. 
        \begin{frame}{Similar triangle form is equivalent to the momentum form}
            \begin{proposition}[Momentum form]\label{prop:r-wapg-momentum-repr}
                Let sequence $(\alpha_k, \rho_k)_{k \ge 0}$ and iterates $(x_k, v_k, y_k)_{k\ge 1}$ be given by the R-WAPG intermediate form.
                Then it has:
                    \item[(i)] For all $k \ge 1$, the iterates $x_{k + 1}, y_{k + 1}$ satisfy:
                        x_{k + 1} &= y_k - L^{-1}\mathcal G_L (y_k),
                        y_{k + 1} &=
                        x_{k + 1} +
                        {\rho_k\alpha_k^2 + \alpha_{k + 1}}(x_{k + 1} - x_k).
                    \item[(ii)] $v_1 = x_1$ if and only if $y_1 = x_1$.
                    \item[(iii)] If in addition $\mu = 0$, then the momentum term in (i) admits a simpler representation:

@@

            A realization of $x_0 \sim \mathcal N(I, \mathbf 0)$ is used for each trial for all algorithms. 
            Each trial results of V-FISTA, M-FISTA (Monotone restarted FISTA) and Free R-WAPG were collected. 
            We plot the medium, minimum, and maximum of $\delta_k$: 

>>>>

% --- Show a title page at the start of each section. 
% \AtBeginSection[]{
%   \begin{frame}
%   \vfill
%   \centering
%   \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
%     \usebeamerfont{title}\insertsectionhead\par%
%   \end{beamercolorbox}
%   \vfill
%   \end{frame}
% }

@@

        \begin{frame}{Other choices for the momentum sequence}
            Generalized by Beck \cite{beck_fast_2009-1} in 2009, called FISTA, it has $(t_k)_{k}$: 

@@

            Chambolle and Dossal in 2015 \cite{chambolle_convergence_2015}, if $t_k = (k + a -1)/a$, then$(x_k)_{k \ge 0}$ has weak convergence in Hilbert space for all $a > 2$.
            \par  
            In Nesterov's book \cite{nesterov_lectures_2018} and Beck's book \cite{beck_first-order_2017} that if in addition, $F:\RR^n \rightarrow \RR$ is $L$ smooth gradient, with Strongly convex with $\mu > 0$, then $(\theta_k)_{k\ge 0}$: 

@@

            It gives $\mathcal O((1 - \sqrt{\mu/L})^k)$ convergence rate for $(F(x_k) - F(x^+))_{k \ge 1}$. 

@@

        \begin{frame}{A major question}
            This is a question that lead our investigation: 
                How much can we relax the choice of the sequence $\theta_k$ and still get an upper bound for the convergence rate? 

@@

@@

            \textbf{Taking this idea further, this is what we did for numerical experiments}: 

@@

                We made a Free R-WAPG algorithm which estimates the strong convexity constant $\mu$ through the Bregman Divergence at two successive iterates while the algorithm is running and, we check the upper bound using our convergence theory. 

@@

            In this section, we highlight key theoretical results that allowed us to unify convergence rate for various variants of FISTA in the literature.

@@

                The algorithm generates a sequence of vector $(y_k, x_{k + 1}, v_{k + 1})_{k \ge 1}$ for all $k\ge 1$ by such that:
                \begin{tcolorbox}\vspace{-1em}

@@

                        L\alpha_k^2 &= (1 - \alpha_k)\gamma_k + \mu \alpha_k, 

@@

                        \hat \gamma_{k + 1} & \defeq L\alpha_k^2,
                        y_k &=

@@

                        g_k &= \mathcal G_L (y_k),

@@

                        v_{k + 1} &=

@@

                        x_{k + 1} &= T_L y_k.

@@

            To use this theorem for the convergence of existing variants of Accelerated Proximal gradient method, we need alternative representations of R-WAPG to fit what commonly appears in the literature. 

@@

        \begin{frame}{Iterates by R-WAPG has other representations}
            \begin{proposition}[Alternative representations of the iterates]\label{prop:wapg-first-equivalent-repr}
                then the iterates can be expressed without $(\gamma_k)_{k \ge1},(\hat \gamma_k)_{k \ge 2}$, and they satisfy for all $k\ge 1$: 

@@

@@

                    & \begin{aligned}

@@

@@

                        &= x_{k + 1} +
                        {\rho_k\alpha_k^2 + \alpha_{k + 1}}(x_{k + 1} - x_k), 
                    \end{aligned}

@@

                    & \begin{aligned}
                        v_{k + 1} 
                        \right) - \frac{1}{L\alpha_{k}}\mathcal G_L (y_k)
                        x_{k + 1} + (\alpha_k^{-1} - 1)(x_{k + 1} - x_k).
                    \end{aligned}

@@

            If $\mu = 0$, then it simplifies:

@@

            A realization of $x_0 \sim \mathcal N(I, \mathbf 0)$ is used for V-FISTA, M-FISTA (Monotone restarted FISTA) and Free R-WAPG. 
            Below are plots of the medium, minimum, and maximum of $\delta_k$: 
